---
title: "Data Modeling"
date: "2025-05-04"
output: 
  html_document:
    df_print: paged
    code_folding: show
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

```{r}
library(lubridate)
library(tidyverse)
library(caret)
library(dplyr)
library(ggthemes)
library(glmnet)
library(MASS)
library(class)
library(pROC)
library(randomForest)
library(ggplot2)
library(corrplot)
library(GGally)
library(rpart)
library(rpart.plot)
library(xgboost)
library(gbm)
library(doParallel)
library(survival)
library(survminer)
boa_data <- read.csv("data/cleaned_data/cleaned_data.csv")
boa_data$closed <- as.factor(boa_data$closed)
```

```{r}
# Recode region: 1 - rural, 2 - micropolitan, 3 - metropolitan
boa_data <- boa_data %>%
  mutate(region_code = case_when(
    is_metropolitan == 1 ~ 3,
    is_micropolitan == 1 ~ 2,
    TRUE ~ 1
  ))
```


## I. Data Exploration

We want to take a look at the data first.

```{r}
# Number of BoA bank overtime
branches_year <- c(5088, 4855, 4752, 4601,
                  4474, 4336, 4253, 4215,
                  3905, 3811, 3743)
year <- 2014:2024

branch_data <- data.frame(year, branches_year)

ggplot(branch_data, aes(x = year, y = branches_year)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "darkblue", size = 2) +
  labs(x = "Year",
       y = "Number of Branches") +
  scale_x_continuous(breaks = 2014:2024) +
  theme_minimal()
```


```{r}
glimpse(boa_data)
```

Now, we want to see within 10 year, how many BoA branches have closed and how many are still opened.

```{r}
boa_data %>%
  mutate(
    region = recode(factor(region_code),
                    `1` = "Rural",
                    `2` = "Micropolitan",
                    `3` = "Metropolitan"),
    branch_status = recode(factor(closed),
                           `0` = "Opened",
                           `1` = "Closed")
  ) %>%
  group_by(region, branch_status) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = region, y = count, fill = branch_status)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), color = "black") +
  geom_text(aes(label = count),
            position = position_dodge(width = 0.9),
            vjust = -0.3, size = 4) +
  labs(
    x = "Region Type",
    y = "Number of Branches",
    fill = "Branch Status"
  ) +
  theme_minimal() + 
  theme(
    legend.position = "top",
    plot.title = element_text(size = 14, face = "bold")
  ) 
```

```{r}
boa_data %>%
  dplyr::select(closed,
         deposits,
         estimate_household_mean_income,
         estimate_household_median_income,
         network_equity) %>%
  pivot_longer(-closed, names_to = "Variable", values_to = "Value") %>%
  mutate(
    Variable = recode(Variable,
      deposits = "Deposits",
      estimate_household_mean_income = "Estimate Household Mean Income",
      estimate_household_median_income = "Estimate Household Median Income",
      network_equity = "Network Equity"
    )
  ) %>%
  ggplot(aes(x = factor(closed), y = Value, fill = factor(closed))) +
  geom_boxplot(outlier.size = 0.8, outlier.alpha = 0.4) +
  facet_wrap(~ Variable, scales = "free_y") +
  scale_y_log10() + 
  labs(
    title = "Boxplots of Key Variables by Closure Status",
    x = "Branch Closure Status (0 = Opened, 1 = Closed)",
    y = "Value (Log 10)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 14, face = "bold")
  )
```


## II. Statistical Analysis

```{r}
# Clean the deposit change column
boa_data$deposit_change <- as.numeric(boa_data$deposit_change)
boa_data$deposit_change[boa_data$deposit_change %in% c(0, -1, Inf, -Inf)] <- NA
```

```{r}
df_cleaned <- boa_data %>%
  dplyr::select(closed,
                combined_stat_area,
                deposits,
                branch_share,
                market_share,
                network_equity,
                deposit_change,
                nearby_boa_branches,
                total_nearby_competitor_branches,
                total_nearby_competitor_deposits,
                total_nearby_boa_deposits,
                estimate_household_median_income,
                estimate_household_mean_income,
                region_code) %>%
  filter(if_all(everything(), ~ !is.na(.) & !is.infinite(.)))
```

Note: 46 of the data points are removed due to the data either being infinite or missing.

```{r}
# Select only numeric columns for correlation
numeric_vars <- df_cleaned %>%
  dplyr::select(deposits,
                branch_share,
                market_share,
                network_equity,
                total_nearby_competitor_branches,
                total_nearby_competitor_deposits,
                estimate_household_median_income,
                estimate_household_mean_income)

cor_matrix <- cor(numeric_vars, use = "complete.obs")

corrplot(cor_matrix,
         method = "color",          
         type = "upper",            
         addCoef.col = "black",     
         number.cex = 0.7,          
         tl.col = "black",          
         tl.srt = 45,               
         tl.cex = 0.9,              
         col = colorRampPalette(c("blue", "white", "red"))(200),  
         mar = c(0, 0, 1, 0)
)

```

### 1. Supervised Learning

We will now split the data into training and testing data (70-30 split).

```{r}
set.seed(123)
train_index <- createDataPartition(df_cleaned$closed, p = 0.7, list = FALSE)
train_data <- df_cleaned[train_index, ]
test_data <- df_cleaned[-train_index, ]
```

#### 1.1. Decision Tree

We will begin by modeling a simple decision tree model.

```{r}
# Fit decision tree
tree_model <- rpart(closed ~ ., 
                    data = train_data, 
                    method = "class", 
                    control = rpart.control(minsplit = 10, cp = 0.01))
```

```{r}
# Predict on test set
pred_class <- predict(tree_model, newdata = test_data, type = "class")

# Confusion matrix and accuracy
conf_mat <- confusionMatrix(pred_class, test_data$closed, positive = "1")
print(conf_mat)
```

```{r}
# Plot the tree
rpart.plot(tree_model,
           type = 2,
           extra = 106,
           fallen.leaves = TRUE,
           box.palette = "Blues")
```

We can see that most of the decisions being made in the tree are related to `deposit_change` and `deposit`. This suggests that the tree is biased towards these two features.

#### 1.2. Boosting

We will now try XGBoost. We will do cross validation to tune the parameters to optimize the boosting model. The main metric we will use is Sensitivity, to ensure we catch as many true branch closures as possible. In addition, we also want to avoid Type II error - situation when the branch is predicted to be active when it is actually closed.

```{r}
# Remove target column and convert predictors to matrix
X_train <- model.matrix(closed ~ . - 1, data = train_data)
y_train <- as.numeric(as.character(train_data$closed))

X_test <- model.matrix(closed ~ . - 1, data = test_data)
y_test <- as.numeric(as.character(test_data$closed))
```

```{r}
train_data$closed <- as.numeric(as.character(train_data$closed))

hyperparam_grid <- expand.grid(
  nrounds = seq(from = 200, to = 500, by = 100),
  eta = c(0.01, 0.1),
  max_depth = c(5, 6, 7),
  gamma = c(0, 1),
  colsample_bytree = c(0.5, 0.75, 1.0),
  min_child_weight = 1,
  subsample = 1
)

num_cores <- parallel::detectCores() - 1
cl <- makeCluster(num_cores)
registerDoParallel(cl)
```

```{r}
tune_control <- trainControl(
  method = "cv", 
  number = 10, 
  verboseIter = FALSE, 
  allowParallel = TRUE
)

xgb_cv <- train(
  closed ~ .,
  data = train_data,
  method = "xgbTree",
  trControl = tune_control,
  metric = "logLoss",
  tuneGrid = hyperparam_grid
)

stopCluster(cl)
registerDoSEQ()
```

After tuning the model, we will use the optimal parameters to fit the final XGBoost model.

```{r}
final_model <- xgboost(data = as.matrix(X_train),
                       label = y_train,
                       booster = "gbtree",
                       objective = "binary:logistic",
                       nrounds = xgb_cv$bestTune$nrounds,
                       max_depth=xgb_cv$bestTune$max_depth,
                       colsample_bytree=xgb_cv$bestTune$colsample_bytree,
                       min_child_weight=xgb_cv$bestTune$min_child_weight,
                       subsample=xgb_cv$bestTune$subsample,
                       eta=xgb_cv$bestTune$eta,
                       gamma=xgb_cv$bestTune$gamma,
                       verbose = 0)
```

```{r}
# Get predicted probabilities
y_prob <- predict(final_model, as.matrix(X_test), type = "response")

# Calculate ROC curve
roc_obj <- roc(test_data$closed, y_prob)
best_threshold <- coords(roc_obj, "best", best.method = "youden")$threshold

# Convert probabilities to classes using optimal threshold
pred_class <- ifelse(y_prob > best_threshold, 1, 0)

# Evaluate
confusionMatrix(factor(pred_class), factor(test_data$closed), positive = "1")
```


```{r}
importance_matrix <- xgb.importance(colnames(X_train), model = final_model)
xgb.plot.importance(importance_matrix, rel_to_first = TRUE, xlab = "Relative importance")
```

### 2. Survival Analysis

#### 2.1. Kaplan-Meier Survival Curve

```{r}
# Step 1: Convert date
boa_data$established_date <- mdy(boa_data$established_date)

# Step 2: Extract year from date
boa_data$established_year <- year(boa_data$established_date)

boa_data_filtered <- boa_data %>%
  filter(established_year < 2024) %>%
  mutate(
    entry_time = pmax(0, 2014 - established_year),
    exit_time  = 2024 - established_year,  # same as duration
    closed = as.numeric(as.character(closed))
  ) %>%
  filter(exit_time > entry_time)

boa_data_filtered <- boa_data_filtered %>%
  dplyr::select(closed,
                exit_time,
                combined_stat_area,
                deposits,
                branch_share,
                market_share,
                network_equity,
                deposit_change,
                nearby_boa_branches,
                total_nearby_competitor_branches,
                total_nearby_competitor_deposits,
                total_nearby_boa_deposits,
                estimate_household_median_income,
                estimate_household_mean_income,
                region_code,
                entry_time
                ) %>%
  filter(if_all(everything(), ~ !is.na(.) & !is.infinite(.)))
```


```{r}
surv_obj <- Surv(time = boa_data_filtered$entry_time,
                 time2 = boa_data_filtered$exit_time,
                 event = boa_data_filtered$closed)

# Kaplan-Meier
km_fit <- survfit(surv_obj ~ 1)

# Plot
plot(km_fit,
     xlab = "Year",
     ylab = "Survival Probability",
     col = "steelblue")
```

#### 2.2. Cox PH Model: Multiple Predictors

```{r}
surv_obj <- Surv(time = boa_data_filtered$entry_time,
                 time2 = boa_data_filtered$exit_time,
                 event = boa_data_filtered$closed)

# Fit Cox PH model with multiple predictors
cox_full <- coxph(surv_obj ~ deposits + deposit_change + 
                    network_equity + estimate_household_median_income + region_code,
                  data = boa_data_filtered)

# Display the summary
summary(cox_full) 
```

#### 2.3. Model Assumption Checking

```{r}
zph <- cox.zph(cox_full)
print(zph)
```

Deposits and deposit_change violates the PH assumption as its effect may change over time. One way to handle that is by interacting deposits with a function of time (like log(t)):

```{r}
cox_time_interact <- coxph(
  Surv(entry_time, exit_time, closed) ~ 
    tt(deposits) + tt(deposit_change) +
    network_equity + estimate_household_median_income + region_code,
  data = boa_data_filtered,
  tt = function(x, t, ...) x * log(t)
)
summary(cox_time_interact) 
```

Before interpreting the results of the Cox Proportional Hazards Model, we examined its key assumptions. 

- First, the model assumes independent observations — that is, branch closures are not correlated due to shared group membership. While we partially account for this by including region_code as a predictor, spatial clustering within regions may still introduce some dependency. 

- Second, we assume non-informative censoring, meaning the reason a branch is still open (censored) is unrelated to its future risk of closure. This cannot be tested directly but is partially addressed by including key predictors like deposit size, income, and network competition that influence both closure and censoring. 

- Finally, the model assumes proportional hazards, meaning each covariate’s effect on closure risk is constant over time. We formally tested this with the cox.zph() function, which showed that this assumption was violated — especially for the deposits variable. 

To correct for this, we implemented a time-dependent Cox model using the tt() function, allowing deposit effects to vary with log-transformed time. This correction ensures our final model better captures the dynamic risk structure over a branch’s lifespan.

To validate our time-interaction Cox model, we conducted 5-fold cross-validation and computed the concordance index for each fold.

```{r}
folds <- createFolds(boa_data_filtered$closed, k = 5, list = TRUE)
c_indices <- numeric(length(folds))

for (i in seq_along(folds)) {
  train_idx <- setdiff(seq_len(nrow(boa_data_filtered)), folds[[i]])
  train_data <- boa_data_filtered[train_idx, ]
  test_data  <- boa_data_filtered[folds[[i]], ]

  # Fit time-dependent Cox model with left truncation
  model <- coxph(Surv(entry_time, exit_time, closed) ~ 
                   tt(deposits) + tt(deposit_change) +
                   network_equity + estimate_household_median_income + region_code,
                 data = train_data,
                 tt = function(x, t, ...) x * log(t))

  # Predict risk scores on test set
  pred <- predict(model, newdata = test_data, type = "lp")

  # Survival object for test set
  surv_test <- Surv(test_data$entry_time, test_data$exit_time, test_data$closed)

  # Concordance index
  c_indices[i] <- survConcordance(surv_test ~ pred)$concordance
}

# Report average C-index
mean_c_index <- mean(c_indices)
print(paste("Average C-index across folds:", round(mean_c_index, 4)))
```

The average C-index across folds was 0.6664, suggesting moderate predictive ability for forecasting branch closure timing. The C-index (concordance index) is analogous to AUC for survival models. 0.663 suggests the model correctly ranks pairs of branch closure risks 66.3% of the time, which is decent for real-world economic data.

